\malg: a novel interest management algorithm for distributed simulations of MMOGs

### 2008-05-15 textos/dsrt/engaoi.tex
last

% :: OLD ABSTRACT ::
Traditionally, a central server is utilized to provide support to MMOGs (massively multiplayer online games), where the number of participants is in the order of tens of thousands. Much work has been done trying to create a fully peer-to-peer model to support this kind of application, in order to minimize the maintenance cost of its infrastructure, but critical questions remain. Examples of the problems relative to peer-to-peer MMOG support systems are: vulnerability to cheating, overload of the upload links of the peers and difficulty to maintain consistency of the simulation among the participants. In this work, it is proposed the utilization of geographically distributed lower-cost nodes, working as a distributed game server. The distribution model and some related works are also presented. To address the communication cost imposed to the servers, we specify the \malg\ algorithm, which is a novel refinement of the area of interest technique, significantly reducing the necessary bandwidth. Simulations have been made with ns-2 and their results demonstrate that our approach achieves the least bandwidth utilization, with a 33.10\% maximum traffic reduction and 33.58\% average traffic reduction, when compared to other area of interest algorithms.

@@

Recently, some alternatives to the central server approach have been researched. One of them is to distribute among the own participants the game simulation and the task to broadcast the state updates when they perform actions. Communication between them is in a peer-to-peer manner, forming a decentralized network \cite{schiele2007rpp}. This approach would be the ideal if it did not have not some critical inherent drawbacks. For example, as the player computers participate on the simulation, they need to agree on the resulting state of the game match. If this agreement does not happen, some inconsistency on the game state among the players may occur.

There is another question, concerning the number of transmissions that each peer needs to execute. In the client-server model, each client only needs to send his actions to the server, which processes them and broadcasts the resulting game state to every player. In the peer-to-peer model, each peer becomes responsible for processing its own player's actions and send the resulting state to everyone else to whom that state change might be relevant. The problem is that we cannot assume that every peer has enough bandwidth to do this. Besides, without a central server, the game becomes dependent on the peers' simulation, which can be corrupted in order to result in an invalid state which benefits a player, or even invalidate the whole game session.

@@

This paper is organized as follows: in section \ref{relatedwork}, some related works with proposed distribution models are presented; in section \ref{def}, some necessary definitions are made; in section \ref{model}, we propose a distribution model to be used as basis of our optimization; in section \ref{aoi}, we review some existing interest management algorithms; in section \ref{gaoi}, we explain the principle of using variable update intervals; in section \ref{csmooth}, we present in detail the \malg\ algorithm; in sections \ref{sim} and \ref{results}, the simulation and its results, respectively, are described and in section \ref{conclusion}, we make our final considerations.

@@

Another work focused on the peer-to-peer model \cite{iimura2004zfg} uses an approach similar to the one presented in \cite{schiele2007rpp}, but suggests, for each region of the virtual environment, the creation of a ``zoned federation'' formed by peers chosen among the participants in that region. Since different nodes will work together, and thus will need to agree in order to proceed with the game, the simulation becomes more trustable. However, the risk of the chosen peers commit collusion cheating \cite{yan2005scc} is not eliminated. Besides, the agreement itself between the peers - that provides increased reliability - creates a fair amount of extra traffic between the participants of the federation, possibly delaying each simulation step.

@@

A key problem in the peer-to-peer architectures, in what concerns the use of interest management, is that some peers are responsible for part of the simulation and for decide to whom each state update interests. Assuming that there are only trustable players, the interest management technique may be useful. However, supposing that a player is malicious, he could hack his client software and avoid sending state updates to certain players, who would be in disadvantage. Consequently, the distributed server model \cite{assiotis2006dam, ng2002msa, fiedler2002cam, rieche2007ppb} is considered more suitable to use the interest management technique.

An example of this kind of model is described in \cite{assiotis2006dam}, where a distributed architecture for MMOGs is proposed. It is also based on the division of the virtual environment into regions, each one with a server node assigned to it. The player whose avatar is situated in a region of the game world should connect to the server responsible for that region. This way, each server would group different players, based on their locality on the virtual space. To achieve consistency between the different server nodes performing the simulation, it is used the concept of locks. When a server node needs to alter the state of some entity of the game, it first needs to obtain exclusive access to that entity. To do that, it negotiates with the other server nodes which might also need to make some change in the same object, and then obtains the lock. When it finishes changing the state, it releases the lock and notifies the other servers

@@

Regarding the interest management technique, works such as \cite{morse1996iml, rak1996egb, zou2001egt, morgan2005imm} and \cite{minson2005aim} may be cited. In \cite{rak1996egb}, it is proposed an interest management scheme based on a virtual environment divided into a grid of cells. Each cell is assigned to a multicast group. Each participant of the simulation subscribes then to the multicast group of the cell where he is, as well as of the neighboring cells if they are within his field of view. Each participant then sends his state updates to the multicast group of the region where he is.

@@

An interest management scheme using a message oriented middleware is presented in \cite{morgan2005imm}. This scheme predicts what will be of interest to a participant in the future, based on his current position and velocity vector in the virtual environment. Thus, each participant starts receiving state updates from entities that are not yet within his field of view, but probably will be in the near future, making their status available as soon as he can see them.

@@

In \cite{morse1996iml}, it is made a survey of systems that use the interest management technique, describing the purpose, scope and what criteria are used by the management scheme employed by each one of them. It is then presented a taxonomy of such systems, classifying them according to: communication model, filtering focus and domain of responsibility. The model may be \textit{unicast}, \textit{multicast} or \textit{broadcast}. The filtering focus refers to what characteristics are observed for each object to perform the filtering: they may be \textit{intrinsic}, such as the values of the object's attributes (e.g. the exact coordinates of its location), or \textit{extrinsic}, such as which multicast group it is subscribed to. Finally, the domain of responsibility assigned to an interest manager, which checks for whom each state update is relevant, can be \textit{static} or \textit{dynamic}. For instance, if an interest manager is assigned to a fixed area of the virtual environment, its domain of responsibility is static, but if he controls an area that may increase or decrease in size, its domain of responsibility is dynamic.

@@

%Será utilizado o termo cliente para referir-se ao computador utilizado por cada jogador para conectar-se a um dos servidores do jogo, assim como o termo servidor fará referência a cada nodo integrante do sistema distribuído que estará servindo o jogo.
It is necessary to describe the network support model on which the proposed interest management algorithm is intended to be utilized. Throughout the text, some terms will be used, and their definitions are given here:

@@

\textbf{Avatar} is the player's representation in the virtual environment. Through his avatar, a player interacts with the game world and with the other players. Examples of avatars are the characters in MMORPGs (massively multiplayer online role-playing games), like World of Warcraft.

@@

\textbf{State} is the set of properties that can be observed in the various entities of the game. The overall state of the simulated world is composed of individual states of all the entities present in it.

@@

The \textbf{border} between two regions is the line that divides the areas that these regions occupy. When an avatar is located near a border, the server responsible for the region beyond this border is notified about the presence of that avatar by the server where it is.

@@

This work is part of the P2PSE \cite{crippa2007pps} project, which is currently based on a peer-to-peer distribution, but still looking for an ideal model. Hence, we propose here an alternative that consists of a virtual environment partitioned into regions - each managed by a different server - on which the \malg\ algorithm is based. The regions are contiguous, exploring the locality of the players' avatars. Thus, avatars next to each other will probably be located in the same region and, therefore, their clients will tend to be connected to the same server, so that their interaction is faster (Figure \ref{partitions}). Situations in which two players interacting with each other are connected to different servers imply a higher bandwidth usage, as it is required some sort of negotiation between the servers to which the different players are connected, in order to maintain the states of the simulation identical in both servers. Also, the hop count of each message between the clients of these players is increased, since there is one more intermediary.

@@

One issue that relates to that type of partitioning of the virtual environment concerns the borders between regions. If an avatar is close to the border between two regions, it will be necessary some exchange of information between the servers which manage them. This information consists of state updates of the entities that are interacting with each other despite being located in different regions. For example, let $S_i$ be the server responsible for the region $R_i$ where the avatar of the client $C_i$ is located and let $S_j$ be the server responsible for another region, $R_j$, where the avatar of the client $C_j$ is situated. When the avatar of $C_i$ gets close to the border with $R_j$, $S_i$ sends to $S_j$ a message notifying it about the presence of that avatar near its region's border. If the avatar of $C_j$ also gets close to the border with $R_i$ and to the avatar of $C_i$, $S_j$ sends another message, notifying $S_i$. Then, $C_i$ and $C_j$ start exchanging state updates of their avatars through the servers.

Regarding the simulation of the actions performed by players whose avatars are located in different regions, it must be decided how it will be performed. As the focus of this work is not the simulation itself, but the bandwidth usage optimization through a new interest management algorithm, it was decided that the simulation will be performed by the server to which the client of that player is connected. Thus, if the player whose avatar is in region $R_i$ performs an action near the border involving entities in $R_j$, it is the server $S_i$ who will decide the outcome of these actions, forwarding only the already calculated new state to $S_j$.

This way, the details of this mechanism will not result in relevant changes to the interest management. When a player $P$, whose avatar is near the border of a region, performs actions whose results must be broadcasted to players with their avatars in other regions, the server $S$, responsible for $P$, simulates his actions, calculates the resulting state and simply sends it to the neighbor server, as if it was sending to its own clients. Similarly, when $S$ receives the resulting state of the action of a player who is connected to a neighbor server, it broadcasts the state to the clients connected to it as if a player in its own region had executed that action.

@@

\section{Interest Management Algorithms}
\label{aoi}

@@

In order to provide an identical sense of the environment among the players, each one of them must maintain a local copy of the entities' states, which must be the same for everyone. The simplest way to do this is to broadcast the states of all entities to all clients. The problem of this approach is that it generates a heavy traffic between the servers and clients, preventing the game to scale well, as the number of participants increase. To save bandwidth, both of the players, as of the servers that intermediate them, a technique known as interest management is employed. This technique reduces the number of updates that the players will receive - and send, in the case of a peer-to-peer architecture.

In short, the interest management technique works as follows: for every state change of each entity, it is calculated to whom it will be important. For example, if an avatar is miles away from another one in the virtual environment, it is most likely that their state alterations are irrelevant to each other. Thus, it is not necessary for them to exchange information on their state. This principle - locality - is used as the main criterion in the interest management algorithms.

The algorithms described in the following sections are mainly based on the euclidean distance between each avatar and all other entities in the virtual environment. This could create a scalability problem due to the processing of the distances, but a distributed architecture is assumed, where the processing can and should be parallelized. In the distribution model defined previously, each server controls a region of the map. Therefore, each one of them manages only a subset of the entities of the game, checking only the distances between each pair of them, in addition to the entities that are in a neighboring region, close to its border.

@@

The simplest way to execute interest management is to define a circular area, whose center is determined by the coordinates of the avatar's location in the virtual environment. After that, the euclidean distance between the avatar and every other entity in the game world is calculated. Let $A$ be an avatar, whose area of interest is a circle of radius $rad$. If an entity $E$ is at a distance less than $rad$ from $A$, then its state updates will be relevant to $A$. $A$ will not receive state updates from entities which are at a distance greater than $rad$. Figure \ref{fcirc} illustrates this type of area of interest.

@@

Before sending the state of an entity to a client, the last transmission time is checked. The next transmission is then scheduled to occur after a certain interval. If the relevance of that state is 1, the normal update interval will be used. If it is less than 1, the normal interval is divided by the relevance. For example, let the normal update interval of a game be \mbox{200 ms}. If the avatar $A_i$, which just sent a state update packet to $A_j$, is at a distance from $A_i$ such that its relevance is 0.5, the next transmission will only happen after an interval of 200 divided by 0.5. So, the player controlling $A_j$ will only receive an update of $A_i$ after 400 ms. Despite this interval is still less than a half second, it represents a reduction of the state update frequency of $A_i$ in 50\%. As they are at a greater distance from one another, and the interval was increased only by 200 ms, this variation will probably be imperceptible by the player who controls $A_j$.

@@

A simple form of using various update intervals based on the relevance value would be utilizing the circular area of interest. To obtain the relevance of an entity relative to an avatar, the value may be set to 1 when they are in the same position and gradually reduce it as the distance between them increases, until it reaches 0 and stops decreasing no matter how farther the entity goes. This is a way that, although simple, has shown a significant reduction in traffic between clients and servers. In Figure \ref{graded_circle}, it is illustrated how this area of interest would be, with entities' state update frequency attenuation for a given avatar.

@@

In the previous sections, we have hinted that filtering based on an avatar field of view of 180 degrees, for instance, instead of using a full circle around it could result in significant reduction in update traffic from the server to the avatar player's machine. However, that would bring an important drawback, which is the large but temporary inconsistencies that would ensue whenever a player would turn around his avatars quickly to see what remotely-controlled objects are behind it. We have also reminded that reducing update frequencies for objects that are farther away from an avatar is a good optimization, especially for large virtual worlds where it may be on an open field with potentially hundreds of other avatars in its sighting distance. In this case, the player will not benefit at all from receiving updates at maximum frequency from objects that are sufficiently distant.
%from using the maximum update frequency that is granted to objects that are at maximum closeness to the player being updated.

@@

The resulting area of interest then takes the shape of a circular sector, which represents the player's field of view. The origin of this circular sector is also the center of a smaller circle, which defines the area close to the avatar. In the close area, all entities have a relevance value of 1, so their update frequency to that avatar is set to the maximum value. This way, the most up-to-date game state becomes available to the player in the area around his avatar, improving the interaction with entities next to it. Even if some of them are outside the player's field of view, he will be able to see them if his avatar turns rapidly to their direction. Figure \ref{fcsmooth} illustrates the area of interest which has just been described.

@@

As for entities that are outside the area nearby, but still within the avatar's field of view, their relevance must be calculated. It is proposed that the relevance of each entity decline gradually as the distance between it and the avatar in question increases. The farther they are, the less frequent will be their state updates. This is possible because even if the interval update is doubled, it will most likely still be a fraction of a second, which is hardly noticeable for a player whose avatar is located at a great distance from the entity in question. Moreover, short delays between the arrival of state updates can be easily masked by extrapolation techniques, such as dead-reckoning \cite{smed2002rna}. The algorithm \ref{smalg} defines the operation of our interest management technique.

@@

Ns-2 simulator \cite{mccanne:nsn} was used to simulate and compare the interest management algorithms described in this paper. This simulator allows the user to create code of the specific application which will be simulated. In our case, it was simulated a server managing a certain region where there was an avatar whose client should receive state updates of the other avatars present in that region.
%In our case, it was simulated a server which should send state updates to the client responsible for one of the avatars in the region managed by that server.
Based on their locations and on the chosen interest management algorithm, the server decided which other avatars had a relevant state to the client in question.

Through the simulations, it was obtained the server upload bandwidth usage for one client. It was not considered necessary to simulate every client connected to the server simultaneously because every one of them had the same behavior. To find the total upload bandwidth usage, we just multiply the value found in the simulation by the number of clients connected to that server. Also, there was no point in simulating or measuring the download bandwidth usage of the server, since it would keep receiving the players' actions at the same rate no matter what interest management algorithm was used.

@@

Some works, such as \cite{yu2007nas}, \cite{kim2005tcm} and \cite{svoboda2007taa}, analyse the network traffic generated by large scale games. Based on these analysis, and adopting a conservative posture, the following parameters have been decided to be used in the simulations:

@@

The results were collected as follows: to measure the average upload bandwidth usage by the server, the sizes of all packets sent in the session were summed up and then divided by the session duration; to determine the maximum usage, it was measured how many bytes had been sent each second and the highest value was selected.

In tables \ref{tab_max} and \ref{tab_avg}, the collected maximum and average upload bandwidth usage data are presented for each number of simulated avatars and interest management algorithm - circular area based (\circleaoi), circular area based with attenuated update frequency (\circleat), field of view based (\fov) and \malg\ algorithm - and it is also show how much would be the upload bandwidth usage if no technique at all was employed (\noaoi). %The values presented are in bytes per second. 
In Figure \ref{fig_max} and Figure \ref{fig_avg}, the corresponding graphics are given.

@@

Regarding the proposed algorithm, \malg, it has been obtained an average upload bandwidth usage reduction of 63.51\% and 33.58\%, compared to the circular area of interest and the field of view based algorithms, respectively. The maximum upload bandwidth usage was also reduced, decreasing 52.03\% and 33.10\%, compared to the same algorithms. Table \ref{tab_summary} shows the maximum and average bandwidth saving percentage of the \malg\ algorithm compared to the other techniques, for average and maximum usage. For example, the maximum bandwidth utilization with \malg\ is 24.81\% less than with a circular area of interest based algorithm with update frequency attenuation.

>>>>

% :: OLD ABSTRACT ::
Traditionally, a central server is utilized to provide support to MMOGs (massively multiplayer online games), where the number of participants is in the order of tens of thousands. Much work has been done trying to create a fully peer-to-peer model to support this kind of application, in order to minimize the maintenance cost of its infrastructure, but critical questions remain. Examples of the problems relative to peer-to-peer MMOG support systems are: vulnerability to cheating, overload of the upload links of the peers and difficulty to maintain consistency of the simulation among the participants. In this work, we propose the utilization of geographically distributed lower-cost nodes, working as a distributed game server. The distribution model and some related works are also presented. To address the communication cost imposed to the servers, we specify the \malg\ algorithm, which is a novel refinement of the interest management technique, significantly reducing the necessary bandwidth. Simulations have been made with ns-2 and their results demonstrate that our approach achieves the least bandwidth utilization, with a 33.10\% maximum traffic reduction and 33.58\% average traffic reduction, when compared to other area of interest algorithms.

@@

Recently, some alternatives to the central server approach have been researched. One of them is to distribute among the own participants the game simulation and the task to broadcast the state updates when they perform actions. The communication between them is peer-to-peer, forming a decentralized network \cite{schiele2007rpp}. This approach would be the ideal if it did not have some critical inherent drawbacks. For example, as the player computers participate on the simulation, they need to agree on the resulting state of the game match. If this agreement does not happen, some inconsistency on the game state among the players may occur.

There is another question, concerning the number of transmissions that each peer needs to execute. In the client-server model, each player only needs to send his actions to the server, which processes them and broadcasts the resulting game state to every player. In the peer-to-peer model, each peer becomes responsible for processing its own player's actions and send the resulting state to everyone else to whom that state change is relevant. The problem is that we cannot assume that every peer has enough bandwidth to do this. Besides, without a central server, the game becomes dependent on the peers' simulation, which may be corrupted in order to result in an invalid state that benefits a player, or even invalidate the whole game session.

@@

This paper is organized as follows: in section \ref{relatedwork}, some related works with proposed distribution models are presented; in section \ref{def}, some necessary definitions are made; in section \ref{model}, we propose a distribution model to be used as basis of our optimization; in section \ref{aoi}, we review the interest management technique; in section \ref{gaoi}, we explain the principle of using variable update intervals; in section \ref{csmooth}, we present in detail the \malg\ algorithm; in sections \ref{sim} and \ref{results}, the simulation and its results, respectively, are described and in section \ref{conclusion}, we make our final considerations.

@@

Another work focused on the peer-to-peer model \cite{iimura2004zfg} uses an approach similar to the one presented in \cite{schiele2007rpp}, but suggests, for each region of the virtual environment, the creation of a ``zoned federation'' formed by peers chosen among the participants in that region. Since different nodes will work together, and thus will need to agree in order to proceed with the game, the simulation becomes more trustable. However, the risk of the chosen peers commit collusion cheating \cite{yan2005scc} is not eliminated. Besides, the agreement itself between the peers - what provides increased reliability - creates a fair amount of extra traffic between the participants of the federation, possibly delaying each simulation step.

@@

A key problem in the peer-to-peer architectures, in what concerns the use of interest management, is that some peers are responsible for part of the simulation and for deciding to whom each state update interests. Assuming that there are only trustable players, the interest management technique may be useful. However, supposing that a player is malicious, he could hack his client software and avoid sending state updates to certain players, who would be in disadvantage. Consequently, the distributed server model \cite{assiotis2006dam, ng2002msa, fiedler2002cam, rieche2007ppb} is considered more suitable to use the interest management technique.

An example of this kind of model is described in \cite{assiotis2006dam}, where a distributed architecture for MMOGs is proposed. It is also based on the division of the virtual environment into regions, but each one with a server node assigned to it. The player whose avatar is situated in a particular region of the game world should connect to the server responsible for that region. This way, each server would group different players, based on their locality on the virtual space. To achieve consistency between the different server nodes performing the simulation, it is used the concept of locks. When a server node needs to alter the state of some entity of the game, it first needs to obtain exclusive access to that entity. To do that, it negotiates with the other server nodes which might also need to make some change in the same object, and then obtains the lock. When it finishes changing the state, it releases the lock and notifies the other servers

@@

Regarding the interest management technique, works such as \cite{morse1996iml, rak1996egb, zou2001egt, morgan2005imm} and \cite{minson2005aim} may be cited. In \cite{rak1996egb}, it is proposed an interest management scheme based on a virtual environment divided into a grid of cells. Each cell is assigned to a multicast group. Each participant of the simulation subscribes then to the multicast group of the cell where he is, as well as of the neighboring cells if they are within his field of view. Each participant then sends his state updates to the multicast group of the region where he is situated.

@@

An interest management scheme using a message oriented middleware is presented in \cite{morgan2005imm}. This scheme predicts what will be of interest to a participant in the future, based on his current position and velocity vector in the virtual environment. Thus, each participant starts receiving state updates from entities that are not yet within his field of view, but probably will be in the near future, making their status available as soon as they can be seen.

@@

In \cite{morse1996iml}, it is made a survey of systems that use the interest management technique, describing the purpose, scope and what criteria are used by the management scheme employed by each one of them. A taxonomy of such systems is presented, classifying them according to: communication model, filtering focus and domain of responsibility. The model may be \textit{unicast}, \textit{multicast} or \textit{broadcast}. The filtering focus refers to what characteristics are observed for each object to perform the filtering: they may be \textit{intrinsic}, such as the values of the object's attributes (e.g. the exact coordinates of its location), or \textit{extrinsic}, such as which multicast group it is subscribed to. Finally, the domain of responsibility assigned to an interest manager, which checks for whom each state update is relevant, can be \textit{static} or \textit{dynamic}. For instance, if an interest manager is assigned to a fixed area of the virtual environment, its domain of responsibility is static, but if it controls an area that may increase or decrease in size, its domain of responsibility is dynamic.

@@

%Será utilizado o termo cliente para referir-se ao computador utilizado por cada jogador para conectar-se a um dos servidores do jogo, assim como o termo servidor fará referência a cada nodo integrante do sistema distribuído que estará servindo o jogo.
It is necessary to describe the network support model on which the proposed interest management algorithm is intended to be utilized. For that reason, some terms will be used throughout the text, and their definitions are given here:

@@

\textbf{Avatar} is the player's representation in the virtual environment. Through his avatar, a player interacts with the game world and with the other players. Examples of avatars are the characters in MMORPGs (massively multiplayer online role-playing games), like World of Warcraft \cite{worldofwarcraft}.

@@

\textbf{State} is the set of properties that can be observed in the various entities of the game. The overall state of the simulated world is composed of the individual states of all the entities present in it.

@@

The \textbf{border} between two regions is the line that divides the areas that these regions occupy. When an avatar is located near a border, the server responsible for the region beyond this border is notified about its presence by the server of the region where that avatar is situated.

@@

This work is part of the P2PSE \cite{crippa2007pps} project, which is currently based on a peer-to-peer distribution, but still looking for an ideal model. Hence, we propose here an alternative that consists of a virtual environment partitioned into regions - each managed by a different server - on which the \malg\ algorithm is based. The regions are contiguous, exploring the locality of the players' avatars. Thus, avatars next to each other will probably be located in the same region and, therefore, their clients will tend to be connected to the same server, so that their interaction is faster (Figure \ref{partitions}).

One issue that relates to that type of partitioning of the virtual environment concerns the borders between regions. If an avatar is close to the border between two regions, it will be necessary some exchange of information between the servers which manage them. This information consists of state updates of the entities that are interacting with each other despite being located in different regions. Such situations imply a higher bandwidth usage, as it is also required some sort of negotiation between the servers to which the different players are connected, in order to maintain the states of the simulation identical among all of them. Furthermore, the hop count of each message between the clients will be increased, since there are two intermediary servers between them.

@@

For example, let $S_i$ be the server responsible for the region $R_i$ where the avatar of the client $C_i$ is located and let $S_j$ be the server responsible for another region, $R_j$, where the avatar of the client $C_j$ is situated. When the avatar of $C_i$ gets close to the border with $R_j$, $S_i$ sends to $S_j$ a message notifying it about the presence of that avatar near its region's border. If the avatar of $C_j$ also gets close to the border with $R_i$ and to the avatar of $C_i$, $S_j$ sends another message, notifying $S_i$. Then, $C_i$ and $C_j$ start exchanging state updates of their avatars through the servers.

It must then be decided how the simulation of actions performed by players whose avatars are located in different regions will be executed. As the focus of this work is not the simulation itself, but the bandwidth usage optimization through a new interest management algorithm, it was decided that the simulation will simply be performed by the server to which that player is connected. Thus, if the player whose avatar is in region $R_i$ performs an action near the border involving entities in $R_j$, it is the server $S_i$ who will decide the outcome of these actions, forwarding only the already calculated new state to $S_j$. This way, the details of the simulation mechanism will not interfere in the interest management. The server $S_i$, responsible for the player $P_i$, simulates his actions, calculates the resulting state and simply sends it to the neighbor server, $S_j$, as if it was sending to its own clients. Similarly, when $S_j$ receives the resulting state of the action of $P_i$, it forwards the received stated to its clients as if a player in its own region had executed that action.

@@

\section{Interest Management}
\label{aoi}

@@

In order to provide an identical sense of the environment among the players, each one of them must maintain a local copy of the entities' states, which must be the same for everyone. The simplest way to do this is to broadcast the states of all entities to everyone. The problem of this approach is that it generates a heavy traffic between the servers and clients, preventing the game to scale well, as the number of participants increase. To save bandwidth, both of the players, as of the servers that intermediate them, a technique known as interest management is employed. This technique reduces the number of updates that the players will receive - and send, in the case of a peer-to-peer architecture.

In short, the interest management technique works as follows: for every state change of each entity, it is calculated to whom it will be relevant. For example, if an avatar is miles away from another one in the virtual environment, it is most likely that their state alterations are irrelevant to each other. Thus, it is not necessary for them to exchange information on their state. This principle - locality - is used as the main criterion in the interest management algorithms.

The algorithms described in the following sections are mainly based on the euclidean distance between each avatar and every other entity in the virtual environment. This could create a scalability problem due to the calculation of the distances, but a distributed architecture is assumed, where the processing can and should be parallelized. In the distribution model defined previously, each server controls a region of the map. Therefore, each one of them manages only a subset of the entities of the game, checking only the distances between each pair of them, in addition to the entities that are in a neighboring region, close to its border.

@@

A simple way to execute interest management is to define a circular area, whose center is determined by the coordinates of the avatar's location in the virtual environment. After that, the euclidean distance between the avatar and every other entity in the game world is calculated. Let $A$ be an avatar, whose area of interest is a circle of radius $rad$. If an entity $E$ is at a distance less than $rad$ from $A$, then its state updates will be relevant to $A$. $A$ will not receive state updates from entities which are at a distance greater than $rad$. Figure \ref{fcirc} illustrates this type of area of interest.

@@

Before sending the state of an entity to a client, the last transmission time is checked. The next transmission is then scheduled to occur after a certain interval. If the relevance of that state is 1, the normal update interval will be used. If it is less than 1, the normal interval is divided by the relevance. For example, let the normal update interval of a game be \mbox{200 ms}. Also, let $A_i$ be the avatar of a player who just received a state update packet from $A_j$. If $A_j$ is at a distance from $A_i$ such that its relevance is 0.5, the next transmission will only happen after an interval of 200 divided by 0.5. So, the player controlling $A_i$ will only receive another update from $A_j$ after 400 ms. Despite this interval is still less than a half second, it represents a reduction of the state update frequency of $A_j$ in 50\%, to the player of $A_i$. As they are at a greater distance from one another, and the interval was increased only by 200 ms, this variation will probably be imperceptible by the player.

@@

A simple form of using various update intervals based on the relevance value would be utilizing the circular area of interest. To obtain the relevance of an entity relative to an avatar, the value may be set to 1 when they are at the same position and gradually reduce it as the distance between them increases, until it reaches 0 and stops decreasing no matter how farther the entity goes. This is a way that, although simple, has shown a significant reduction in traffic between clients and servers. In Figure \ref{graded_circle}, it is illustrated how this area of interest would be, with entities' state update frequency attenuation for a given avatar.

@@

In the previous sections, we have hinted that filtering based on an avatar's field of view of 180 degrees, for instance, instead of using a full circle around it could result in significant reduction in update traffic from the server to the player's machine. However, that would bring an important drawback, which is the large but temporary inconsistencies that would ensue whenever a player would turn around his avatars quickly to see what remotely-controlled objects are behind it. We have also reminded that reducing update frequencies for objects that are farther away from the observer is a good optimization, especially for large virtual worlds where the avatar may be on an open field with potentially hundreds of other avatars in its sighting distance. In this case, the player will not benefit at all from receiving updates at maximum frequency from objects that are sufficiently distant.
%from using the maximum update frequency that is granted to objects that are at maximum closeness to the player being updated.

@@

The resulting area of interest then takes the shape of a circular sector, which represents the player's field of view. The origin of this circular sector is also the center of a smaller circle, which defines the area close to the avatar. In the close area, all entities have a relevance value of 1, so their update frequency to that avatar is set to the maximum value. This way, the most up-to-date game state becomes available to the player in the area around his avatar, improving the interaction with entities next to it. Even if some of them are outside the player's field of view, he will be able to see them if his avatar turns rapidly in their direction. Figure \ref{fcsmooth} illustrates the area of interest which has just been described.

@@

As for entities that are outside the area nearby, but still within the avatar's field of view, their relevance must be calculated. It is proposed that the relevance of each entity decline gradually as the distance between it and the avatar in question increases. The farther they are, the less frequent will be their state updates. This is possible because even if the interval update is doubled, it will most likely still be a fraction of a second, which is hardly noticeable for a player whose avatar is located at a great distance from that entity. Moreover, short delays between the arrival of state updates can be easily masked by extrapolation techniques, such as dead-reckoning \cite{smed2002rna}. Algorithm \ref{smalg} defines the operation of our interest management technique.

@@

The ns-2 simulator \cite{mccanne:nsn} was used to compare the interest management algorithms described in this paper. This simulator allows the user to create code of the specific application which will be simulated. In our case, it was simulated a server managing a certain region where there was an avatar whose client should receive state updates of the other avatars present in that region.
%In our case, it was simulated a server which should send state updates to the client responsible for one of the avatars in the region managed by that server.
Based on their locations and on the chosen interest management algorithm, the server decided which other avatars had a relevant state to the client in question.

Through the simulations, it was obtained the server upload bandwidth usage for one client. It was not considered necessary to simulate every client connected to the server simultaneously because every one of them had the same behavior. To find the total upload bandwidth usage, we just multiply the value measured with the simulation by the number of clients connected to that server. Also, there was no point in simulating or measuring the download bandwidth usage of the server, since it would keep receiving the players' actions at the same rate no matter what interest management algorithm was used.

@@

Some works, such as \cite{yu2007nas}, \cite{kim2005tcm} and \cite{svoboda2007taa}, analyse the network traffic generated by large scale games. Based on these analyses, and adopting a conservative posture, the following parameters have been decided to be used in the simulations:

@@

The results were collected as follows: to measure the average upload bandwidth usage by the server, the sizes of all packets sent in the session were added up and then divided by the session duration; to determine the maximum usage, it was measured how many bytes had been sent each second and the highest value was selected.

In tables \ref{tab_max} and \ref{tab_avg}, the collected maximum and average upload bandwidth usage data are presented for each number of simulated avatars and interest management algorithm - circular area based (\circleaoi), circular area based with attenuated update frequency (\circleat), field of view based (\fov) and \malg\ algorithm - and it is also shown how much would be the upload bandwidth usage if no technique at all was employed (\noaoi). %The values presented are in bytes per second. 
In Figure \ref{fig_max} and Figure \ref{fig_avg}, the corresponding graphics are given.

@@

Regarding the proposed algorithm, \malg, it has been obtained an average upload bandwidth usage reduction of 63.51\% and 33.58\%, compared to the circular area of interest and the field of view based algorithms, respectively. The maximum upload bandwidth usage was also reduced, decreasing 52.03\% and 33.10\%, compared to the same algorithms. Table \ref{tab_summary} shows the saving percentage of the maximum and average bandwidth usage with the \malg\ algorithm, compared to the other simulated techniques. For example, the maximum bandwidth utilization with \malg\ is 24.81\% less than with a circular area of interest based algorithm with update frequency attenuation.
